---
title: "Assignment 3: Data Exploration"
author: "Jordan Heinlein"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup,  message = FALSE}
#installing the following packages. I have commented out after running the code 
#to avoid issues knitting the document.
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("here")
#install.packages("tidyverse")
#install.packages("lubridate")
#The above packages are installed, packages are now loaded for this session.
library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)
library(lubridate)
#Confirm location of working directory
getwd()
here()
#Importing the data sets by reading in the two data files using relative file 
#locations. Assigning names of "Neonics" and "Litter". Also setting strings as 
#factors as true. 
Neonics <- read.csv(
  file= here('Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'),
  stringsAsFactors = TRUE
)

Litter <- read.csv(
  file= here('Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
  stringsAsFactors = TRUE
)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Understanding the ecotoxicology of neonicotinoids and their effects
on insects is important because, as stated above, this class of insecticide is
widely used to control pests in agricultural settings. However, neoicotinoids 
are known to pose ecological risk to all insects, not just pest species. 
Therefore, the ecotox is important not only from a pure conservation standpoint, 
but also because their use is potentially detrimental to other insects that are 
pollinators and provide critical services to our agricultural systems.
Consequently, it is important to understand the nature and extent to which these
insecticides impact non-pest insects, like pollinators, which may allow
regulators to understand the tradeoffs of neonicotinoid use. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Litter and woody debris that falls to the ground in forests might give 
us information about the dominant types of trees in the system, i.e. broadleaf
versus coniferous trees and ecosystem health. Additionally, according to the 
NEON User Guide, this data can be used to estimate other ecological parameters
like NPP, biomass, and carbon flux (Jones 2017). 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Samples are collected in randomly selected tower plots.
 2.In plots, litter (elevated traps) and woody debris (ground traps) are paired,
 with several pairs per plot in the study area. 
 3.Trap placement within plots is either random or targeted depending upon 
 vegetation factors like %cover, height, and spatial distribution. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}
dim(Neonics)
#The dimensions of the Neonics dataset is 4623 observations by 30 variables. 
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6}
#Using summary command on the Neonics dataset, specifying "Effect" column using
#$. Applying sort() lists the values in the column by occurrence.
sort(summary(Neonics$Effect))

```

> Answer: The most commonly studdied effect is population. This might be
specifically of interest because researchers may way to understand the risk of a 
neonicotinoid causing the collapse of a population of insects. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7}
#Creating a summary of species common name column in the Neonics dataset, and 
#sorting the results using the maxsum argument to return the 7 highest levels 
#of the factor. 
(summary(Neonics$Species.Common.Name,maxsum = 7)

```

> Answer: Not including "other" species (3083), the six most commonly studied 
species in the dataset are: Honey Bee(667), Parasitic Wasp(285), 
Buff Tailed Bumblebee(183), Carniolan Honey Bee (152), Bumble Bee (140), and 
Italian Honeybee (113). Other than the parasitic wasp, they are all bees.
These are likely of interest because they beneficial insects for agricultureare,
either providing pollinator services or potentially parasitism on pest species. 
Therefore understanding the ecotox impacts on these species is economically 
significant.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}
str(Neonics)
```

> Answer: The Conc.1..Author column is a factor and not numeric because some of 
the entries also contain character symbols like '/' and '>', so the data 
are not solely numeric values. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Q9}
#Use ggplot to create a plot using Neonics, specifying geom_freqpoly with 
#publication year as the x axis. Counts are on the y axis. 
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year))

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Q10}
#Same code as above, but now color is set to the categorical variable 
#Test.Location.
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year, color= Test.Location))

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Recently, the lab appears to be the most common test location, 
followed by natural field locations. There are relatively few artificial field 
locations in the dataset. At the beginning of the dataset in the 1980s, lab 
tests were uncommon and increased through time, peaking in the mid 2010s.
During certain periods, natural field tests have been more common than lab 
tests. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}
#Created bar graph of endpoints using ggplot() and geom_bar, setting x axis as 
#the endpoint varaible. 
ggplot(Neonics) +
  geom_bar(aes(x=Endpoint,)) 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer: The two most common end points are NOEL followed by LOEL. NOEL is the 
highest concentration producing effects not significantly different from those
of the control individuals. LOEL is the lowest concentration producing effects
significantly different from that of the controls. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}
#Running the following code, collectDate is a factor, not a date.
class(Litter$collectDate)
#Change collectDate to a date class using lubridate() and ymd format. 
Litter$collectDate <- ymd(Litter$collectDate)
#Confirming the collect.Date is now a date class. 
class(Litter$collectDate)

unique(Litter$collectDate)
#Litter was sampled on August 2nd and 30th 2018. 
```

13. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}
#Generating the unique plot IDs or "levels".
unique(Litter$plotID)
#Generating summary information for the plotID column to compare. 
summary(Litter$plotID)

```

> Answer: The unique() command just returns the different levels the plotID 
varaible can take on in this dataset. Summary provides additional information 
such as the frequency of each plotID in the dataset. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}
ggplot(Litter) +
  geom_bar(aes(x= functionalGroup))

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Q15}
#Boxplot of dryMass by functionalGroup.
ggplot(Litter)+
  geom_boxplot(aes(x=functionalGroup, y= dryMass))
#Violin plot of dryMass by functionalGroup.
ggplot(Litter)+
  geom_violin(aes(x=functionalGroup, y= dryMass))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:The violin plot in not an effective visualization in this case because
not much information is communicated by the density distribution across the 
different functional group categories. The violin plot of this set obscures
useful information shown by the boxplot like the median values, IQR, skew, 
outliers,etc.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Based on the boxplot of dryMass by functionalGroup, needles and the 
mixed group have the highest biomass as shown by the higher medians, IQRs, 
and whiskers.These types are followed by twigs/branches. 

